---
title: "Gallstone Bayesian Analysis"
subtitle: "Stats 115 Final Project"
author: "Olivia Willard, YuNing Chou, Allinn Chen, Mihir Borkar"
format:
  revealjs:
    slide-number: true
    incremental: false
    theme: simple
editor: visual
---

## Topic and Motivation

-   Gallstones are a common condition that can lead to serious complications\
-   Many clinical and body composition variables may influence gallstone risk\
-   It is unclear which predictors are most important

**Goal:** Identify key predictors of gallstones and estimate risk

## Research Question

**Initial Question:** Are gallstone more prevelant among patients with comorbidities compared to those without comorbidities? Can comorbidity predict gallstone?

**Refined Question:** How do clinical and body composition variables influence the probability that a patient has gallstones?

## Dataset

-   Source: UCI Machine Learning Repository
    -   Contains clinical data collected from the Internal Medicine Outpatient Clinic of Ankara VN Medical Park Hospital in Turkey
    -   Contains information on 319 patients, of whom 161 were diagnosed with gallstone disease
    -   Contains 38 variables describing patient health and characteristics

## Data Cleaning

-   Renamed variables for clarity
-   We converted variables to binary:
    -   Gallstone (reserved coding)
        -   0 = does not have gallstones
        -   1 = has gallstones
    -   Comorbidity
        -   0 = no comorbidities
        -   1 = at least one comorbidity

## EDA: Gallstone Outcomes

```{r}
#| echo: false
#| message: false
#| warning: false
#| out-width: 100%
#| fig-align: center

library(ggplot2)
library(tidyverse)

data <- read.csv("data/cleaned_data.csv")
ggplot(data, aes(x = factor(gallstone))) +
  geom_bar() +
  labs(
    title = "Distribution of Gallstone Status",
    x = "Gallstone Status (1 = Gallstones, 0 = No Gallstones)",
    y = "Count"
  )
```

## EDA: Comorbidity vs Gallstones

```{r}
#| echo: false
#| message: false
#| warning: false
#| out-width: 100%
#| fig-align: center


ggplot(data, aes(x = factor(comorbid), fill = factor(gallstone))) +
  geom_bar(position = "fill") +
  labs(
    title = "Comorbidity vs Gallstone Status",
    x = "Comorbidity (0 = none, 1 = at least one)",
    y = "Proportion",
    fill = "Gallstones"
  )
```

## EDA: Correlation Between Variables

```{r}
#| echo: false
#| message: false
#| warning: false
#| out-width: 100%
#| fig-align: center

library(ggcorrplot)
library(bayesrules)
library(rstanarm)
library(bayesrules)
library(tidyverse)
library(rstan)
library(rstanarm)
library(posterior)
library(tidybayes)
library(janitor)
library(broom.mixed)
library(bayesplot)
library(loo)
library(dplyr)


numeric_data <- data |> 
  select(where(is.numeric))

corr <- cor(numeric_data)

ggcorrplot(
  corr,
  lab = FALSE,
  colors = c("blue", "white", "red"),
  tl.cex = 8
)
```

## EDA: Conclusion

-   Comorbidity showed weak predictive value\
-   Other clinical and body composition variables showed stronger relationships

**Decision:**

-   Shift focus from comorbidity → multiple predictors

## Model Choice

::: {style="font-size: 0.7em; line-height: 1.1;"}
We used a **Bayesian Logistic Regression** model

**Why?**

-   Binary outcome: gallstones vs. no gallstones
-   Directly models probabilities
-   Incorporates prior information
-   Provides uncertainty estimates
:::

::: {style="font-size: 0.7em;"}
$$
\text{logit}(p_i) = \beta_0 + \beta_1(\text{vita_d}_i) + \beta_2(\text{bone_mass}_i) + \beta_3(\text{crp}_i) + \beta_4(\text{icw}_i) + \beta_5(\text{hgb}_i)
$$
:::

::: {style="font-size: 0.7em; line-height: 1.1;"}
-   Main predictors selected based on EDA findings: Vitamin D, bone mass, CRP, ICW, and HGB
:::

## Prior Model

::: {style="font-size: 0.6em; line-height: 1.1;"}
```{r}
#| eval: true
#| echo: true
#| fontsize: 0.5em

# fit prior predictive model
prior_model <- stan_glm(
  formula = gallstone ~ vita_d + bone_mass + crp + icw + hgb,
  data = data,
  family = binomial(link = "logit"),
  prior_intercept = normal(
    location = qlogis(0.0605),
    scale = abs(qlogis(0.065) - qlogis(0.056)) / 4,
    autoscale = F
  ),
  prior = normal(location = 0, scale = 2.5, autoscale = T),
  chains = 4, iter = 10000, seed = 84735,
  prior_PD = T, refresh = F, cores = 16
)
```

-   Intercept:
    -   Intercept prior based on baseline gallstone prevalence obtained from research (\~6%)
-   Coefficients:
    -   Weakly informative Normal priors centered at 0 for regression coefficients (0, 2.5)

**Why?**

-   Prevent extreme estimates
-   Still allow data to dominate
:::

## Model Checking and Results

-   Trace plots showed good chain mixing
-   R-hat values were close to 1 (convergence achieved)
-   Density plot shows consistent chains
-   Effective sample sizes were acceptable
-   Posterior predictive checks suggested the model captured the general pattern of the data

**Conclusion:**

-   Model is reliable

## Posterior Model

```{r}
#| eval: true
#| echo: true
#| fontsize: 0.5em

# fitting the posterior model using the observed data
posterior_model <- update(prior_model, prior_PD = FALSE)
```

-   After specifying the priors, we updated the model using the observed data to obtain the posterior distribution.

-   This allowed us to estimate how each predictor was associated with the probability of gallstones while incorporating uncertainty.

## Model Comparison

::: {style="font-size: 0.5em; line-height: 1.1;"}
We have tested with 6 different models containing the different amount of predictors:

-   Model 1 (7 predictors): vita_d + bone_mass + crp + icw + hgb + hfa + gender
    -   Accuracy about 81.82%
-   Model 2 (6 predictors): vita_d + bone_mass + crp + icw + hgb + hfa
    -   Accuracy about 80.25%
-   Model 3 (5 predictors): vita_d + bone_mass + crp + icw + hgb
    -   Accuracy about 79.62%
-   Model 4 (4 predictors): vita_d + bone_mass + crp + hgb
    -   Accuracy about 78.99%
-   Model 5 (3 predictors): vita_d + bone_mass + crp
    -   Accuracy about 79.62%
-   Model 6 (1 predictor): comorbid
    -   Accuracy about 49.53%
:::

## Model Comparison

::: {style="font-size: 0.5em; line-height: 1.1;"}
Here are the CV Classification Accuracy outputs and ELPD comparison:

-   Model 1 (7 predictors): vita_d + bone_mass + crp + icw + hgb + hfa + gender
    -   Sens: \~81.63%, Spec: \~77.59%, Acc: \~79.64%; ELPD_diff: -0.2, SE_diff: 2.3
-   Model 2 (6 predictors): vita_d + bone_mass + crp + icw + hgb + hfa
    -   Sens: \~81.13%, Spec: \~78.24%, Acc: \~79.92%; ELPD_diff: -0.1, SE_diff: 2.0
-   Model 3 (5 predictors): vita_d + bone_mass + crp + icw + hgb
    -   Sens: \~79.28%, Spec: \~77.70%, Acc: \~78.72%; ELPD_diff: 0, SE_diff: 0
-   Model 4 (4 predictors): vita_d + bone_mass + crp + hgb
    -   Sens: \~77.17%, Spec: \~79.25%, Acc: \~78.37%; ELPD_diff: -21.6, SE_diff: 7.6
-   Model 5 (3 predictors): vita_d + bone_mass + crp
    -   Sens: \~74.59%, Spec: \~83.09%, Acc: \~78.69%; ELPD_diff: -21.0, SE_diff: 7.6
-   Model 6 (1 predictor): comorbid
    -   Sens: \~0%, Spec: \~100%, Acc: \~49.52%; ELPD_diff: -256.5, SE_diff: 41.4

**Results:**

-   Models 1, 2, and 3 performed best and performed similarly under cross-validation and LOOCV
-   Model 6 (comorbidity only):
    -   \~49% accuracy → very poor
:::

## Final Model Choice

::: {style="font-size: 0.7em; line-height: 1.1;"}
We selected **Model 3**:

-   Predictors: Vitamin D, bone mass, CRP, ICW, HGB
-   Balanced predictive performance and simplicity
-   Similar performance to larger models with fewer predictors
-   Better sensitivity than the smaller 3-predictor model

Performance:

-   Overall accuracy: about **79–80%**
-   Better at identifying patients without gallstones than patients with gallstones
:::

## Model Checking and Results

The model suggests:

-   Higher **Vitamin D**, **bone mass**, and **HGB** are associated with higher odds of gallstones
-   Higher **CRP** and **ICW** are associated with lower odds of gallstones

Because the posterior intervals did not include 0, these predictors appeared meaningfully related to gallstone status in this dataset.

## Fairness and How Wrong

::: {style="font-size: 0.7em; line-height: 1.1;"}
-   Data is collected by the hospital, without harm to patients
    -   stripped of identifying features therefore anonymized
-   collected for ML to predict gallstones
    -   purpose being early detection of gallstones in patients
-   We don't observe any negative impacts
    -   preemptive measure
    -   notifying patients to get checked
-   Located in Turkey
    -   regional bias for detection accuracy
-   Posterior predictive check slightly off and requires tuning
:::

## Key Takeaways

::: {style="font-size: 0.7em; line-height: 1.1;"}
-   Comorbidity alone was not a strong predictor in this dataset
-   A combination of clinical and body composition variables provided much better prediction
-   Bayesian logistic regression helped us estimate gallstone risk while clearly showing uncertainty
-   Bayesian approach provides:
    -   Probabilities
    -   Uncertainty quantification
    -   Interpretability
-   EDA helped refine the question, and Bayesian modeling helped us answer it in a principled way
:::

## Who Benefits?

-   Researchers studying gallstone risk factors
-   Clinicians interested in screening-related variables (better risk assessment)
-   Patients can receive earlier detection
-   Future work on medical risk prediction using Bayesian models

## Limitations

-   Limited dataset and sample context
-   Possible missing predictors
-   Comorbidity may have been too simplified after binarization
-   Some predictors may be correlated or reflect similar biological information
-   The model slightly underestimated the observed gallstone proportion

## Future Work

If we had unlimited time/resources:

-   Use a larger and more diverse dataset
-   Include richer clinical information
-   Explore more detailed comorbidity measures
-   Compare additional Bayesian models and nonlinear effects
-   Evaluate external validation on a new dataset

## Final Message

-   Bayesian modeling helps us move beyond simple prediction
-   It allows us to understand **uncertainty and probability**
-   Our results highlight important predictors of gallstone risk

## Thank you!
